---
title: "Group1"
output:
  pdf_document: default
  html_document: default
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 5)
```

## Introduction

## Literature Review

## Theory

### Hypothesis

## Exploratory Data Analysis
```{r echo=FALSE, warning=FALSE}
library(tidyverse)
library(psych)
library(ggpubr)
library(corrplot)
library(rpart)
library(rpart.plot)
library(caret)
library(tidyr)

data <- read.csv("Data/wvs_data.csv")
```

```{r echo=FALSE, warning=FALSE}
data <- data %>%
  dplyr::mutate(happiness = dplyr::recode(V10, 
                                   '1' = 4,
                                   '2' = 3,
                                   '3' = 2,
                                   '4' = 1, 
                                   '-2' = NULL),
         
         health = dplyr::recode(V11,
                                '1' = 5, 
                                '2' = 4, 
                                '3' = 3,
                                '4' = 2,
                                '5' = 1, 
                                '-2' = NULL),
         
         finances = dplyr::recode(V59, 
                                  '-2' = NULL,
                                  .default = V59),
         
         satisfaction = dplyr::recode(V23, 
                                  '-2' = NULL,
                                  .default = V23),
         
         freedom = dplyr::recode(V55, 
                                  '-2' = NULL,
                                  .default = V55),
         gone_no_food = dplyr::recode(V188,
                                      '-2' = NULL,
                                      .default = V188),
         gone_no_health = dplyr::recode(V190,
                                        '-2' = NULL,
                                        .default = V190),
         gone_no_income = dplyr::recode(V191,
                                        '-2' = NULL,
                                        .default = V191),
         trust_neighborhood = dplyr::recode(V103,
                                            '-2' = NULL,
                                            .default = V103),
         social_class = dplyr::recode(V236,
                                      '-2' = NULL,
                                      .default = V236))

fa.fit <- fa(data[,c("happiness",
                         "health",
                         "finances", 
                         "satisfaction",
                         "freedom")], 
             nfactors=1)

data$life.quality <- as.numeric(fa.fit$scores)

cols <- c("gone_no_food", "gone_no_health", "gone_no_income", "trust_neighborhood", "social_class", "life.quality")
sub_data <- data %>% select(all_of(cols))
```

### Missing Data Plot
```{r echo=FALSE, warning=FALSE}
na_counts <- colSums(is.na(sub_data[, cols]))

barplot(na_counts,
        main = "Missing Values per Column",
        xlab = "Column",
        ylab = "Count of NAs",
        col = "steelblue",
        las = 2)  # rotate x labels
```

### Data Cleaning
```{r echo=FALSE, warning=FALSE}
sub_data <- sub_data[!is.na(sub_data$life.quality), ]
predictors <- setdiff(cols, "life.quality")

sub_data <- na.omit(sub_data)
sub_data$life.quality <- as.numeric(sub_data$life.quality)
sub_data[, predictors] <- lapply(sub_data[, predictors], as.numeric)
```

## Methodology

### Assumptions

### Testing and Validation
#### Unpruned and Pruned trees comparison
```{r echo=FALSE, warning=FALSE}
set.seed(123)
train_idx <- sample(nrow(sub_data), 0.8 * nrow(sub_data))
train <- sub_data[train_idx, ]
test  <- sub_data[-train_idx, ]

tree_unpruned <- rpart(
  life.quality ~ .,
  data      = train[, c(predictors, "life.quality")],
  method    = "anova",       # use "class" for classification
  control   = rpart.control(cp = 0, minsplit = 2, minbucket = 1)
)
best_cp <- tree_unpruned$cptable[
  which.min(tree_unpruned$cptable[, "xerror"]), "CP"
]

tree_pruned <- prune(tree_unpruned, cp = best_cp)
# â”€â”€ 4. Compare on test set
pred_unp  <- predict(tree_unpruned, test[, predictors])
pred_prun <- predict(tree_pruned,   test[, predictors])
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
```

#### K-Folds cross validation
```{r echo=FALSE, warning=FALSE}
set.seed(123)
# â”€â”€ 1. Set up k-fold CV (k=10 is standard)
ctrl <- trainControl(method = "cv", number = 10)

# â”€â”€ 2. Train with CV
cv_tree <- train(
  life.quality ~ .,
  data      = train[, c(predictors, "life.quality")],
  method    = "rpart",
  trControl = ctrl,
  tuneLength = 10        
)

plot(cv_tree)            

pred_cv <- predict(cv_tree, newdata = test[, predictors])

```

#### Bagging validation:
```{r echo=FALSE, warning=FALSE}
set.seed(123)
n_bags  <- 10          # number of bootstrap samples
n_train <- nrow(train)

# Store each tree
trees <- vector("list", n_bags)

for (i in 1:n_bags) {
  # Bootstrap sample
  idx       <- sample(1:n_train, n_train, replace = TRUE)
  boot_data <- train[idx, c(predictors, "life.quality")]
  
  # Fit tree on bootstrap sample
  trees[[i]] <- rpart(
    life.quality ~ .,
    data    = boot_data,
    method  = "anova",
    control = rpart.control(cp = 0.01)
  )
}

preds_matrix <- sapply(trees, function(tree) {
  predict(tree, newdata = test[, predictors])
})

# Average predictions across all trees (this is bagging)
pred_bag <- rowMeans(preds_matrix)

par(mfrow = c(2, 2))   # show 6 trees
for (i in 1:2) {
  rpart.plot(trees[[i]],
             main = paste("Bootstrap Tree", i),
             cex  = 0.6,
             type = 2)
}
par(mfrow = c(1, 1))
```
Where we try to observe the different decision trees formed by using bootstrap samples of the dataset provided for our interested features.
## Result Analysis

### Feature Importance observed through K-folds validation where the number of folds = 10
```{r echo=FALSE, warning=FALSE}
plot(varImp(cv_tree),
     main = "Feature Importance - 10-Fold cross validation Tree",
     top  = 10)  
```

### The goodness of fit ($R^{2}$) for each model and the root mean squared error for each model
```{r echo=FALSE, warning=FALSE}
r2 <- function(actual, pred) {
  ss_res <- sum((actual - pred)^2)
  ss_tot <- sum((actual - mean(actual))^2)
  1 - (ss_res / ss_tot)
}

results <- data.frame(
  Model = c("Unpruned Tree", "Pruned Tree", "CV Tree", "Bagging"),
  RMSE  = round(c(
    rmse(test$life.quality, pred_unp),
    rmse(test$life.quality, pred_prun),
    rmse(test$life.quality, pred_cv),
    rmse(test$life.quality, pred_bag)
  ), 4),
  R2 = round(c(
    r2(test$life.quality, pred_unp),
    r2(test$life.quality, pred_prun),
    r2(test$life.quality, pred_cv),
    r2(test$life.quality, pred_bag)
  ), 4)
)

# Sort by RMSE ascending (best first)
results <- results[order(results$RMSE), ]

print(results)
```

### Plotting goodness of fit for each feature $R^{2}$ used in the decision tree.
```{r echo=FALSE, warning=FALSE}
# RÂ˛ per feature = correlationÂ˛ between feature and life.quality
feature_r2 <- sapply(predictors, function(var) {
  cor(as.numeric(test[[var]]), test$life.quality, use = "complete.obs")^2
})

# Convert to data frame
feature_r2_df <- data.frame(
  Variable = names(feature_r2),
  R2       = round(feature_r2, 4)
) |> arrange(desc(R2))

ggplot(feature_r2_df, aes(x = reorder(Variable, R2), y = R2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "RÂ˛ per Feature vs Life Quality",
       x     = "Variable",
       y     = "RÂ˛") +
  theme_minimal()
```
## Testing Decision Tree Validity and Reliability

### Correlation matrix
```{r plotting correlation matrix, echo=FALSE, warning=FALSE}
cor_mat <- cor(train[, c(predictors, "life.quality")], use = "complete.obs")
cor_mat
```

```{r echo=FALSE, warning=FALSE}
corrplot(cor_mat, method = "color", addCoef.col = "black", tl.col = "black", tl.srt = 45, title = "Correlation Matrix of QoL Predictors", mar = c(0, 0, 2, 0))
```
#### Observations
- Positive correlation between health, income and food although the relationship itself is quite weak (score of 0.29 with food and score of 0.36 with both health and food).
- All three variables have positive correlations score with each other, indicating that these factors affect each other.
- Very weak positive relationship between trust in neighbourhoods and social class. 

### Skewness Checks

```{r skewness check, echo=FALSE, warning=FALSE}
skew_values <- sapply(train[, c(predictors, "life.quality")], skew, na.rm=TRUE)

skew_values_df <- data.frame(
  skewness = round(skew_values, 3)
) |> arrange(desc(abs(skewness)))

print(skew_values_df)
```

```{r plot skewness distribution for all variables}
# Pivot skew measurement data frame from wide to long
train[, c(predictors, "life.quality")] |> 
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") |> 
  ggplot(aes(x=Value)) + 
  geom_histogram(bins = 30) + facet_wrap(~Variable, scales="free")
```
#### Discussion
- For variables `gone_no_food`, `gone_no_health`, `gone_no_income`, majority of survey responses sit at highest possible score. This also confirms left-skew distribution for `life.quality`. 
- Very few respondents indicated poor quality of life factors (not enough food, bad health and low income). 
- Indicates that ceiling effect might be occurring, a measurement limitation that is occurring in the survey due to a low upper limit.
- Graph distribution of `life.quality` values show a left skew, these match skewness values shown above.
- `social_class` more balanced, values concentrated on response values 2-4

### Detecting model for overfitting
```{r overfitting detection, echo=FALSE, warning=FALSE}
rmse <- function(actual, pred) {
  sqrt(mean((actual - pred)^2))
}

# Make predictions on training set

pred_unp_train <- predict(tree_unpruned, train[, predictors])
pred_prun_train <- predict(tree_unpruned, train[, predictors])
pred_cv_train <- predict(cv_tree, newdata = train[, predictors])
pred_bag_train <- rowMeans(sapply(trees, function(t) predict(t, newdata = train[, predictors])))

results_overfit_check <- data.frame(
  Model = c("Unpruned Tree", "Pruned Tree", "CV Tree", "Bagging"),
  train_RMSE  = round(c(
    rmse(train$life.quality, pred_unp_train),
    rmse(train$life.quality, pred_prun_train),
    rmse(train$life.quality, pred_cv_train),
    rmse(train$life.quality, pred_bag_train)
  ), 4),
  test_RMSE = round(c(
    rmse(test$life.quality, pred_unp),
    rmse(test$life.quality, pred_prun),
    rmse(test$life.quality, pred_cv),
    rmse(test$life.quality, pred_bag)
  ), 4)
)

results_overfit_check$RMSE_gap <- round(results_overfit_check$test_RMSE- results_overfit_check$train_RMSE, 4)
results_overfit_check <- results_overfit_check[order(results_overfit_check$RMSE_gap), ]

print(results_overfit_check)
```
```{r model performance graph, echo=FALSE, warning=FALSE}
results_overfit_check_long <- results_overfit_check |> 
  pivot_longer(cols = c("train_RMSE", "test_RMSE"),
               names_to = "Set",
               values_to = "RMSE")

ggplot(results_overfit_check_long, aes(x=Model, y=RMSE, fill=Set)) + geom_bar(stat="identity", position="dodge") +
  labs(title = "Training vs Test RMSE Model Comparison")
```

#### Discussion
- Tree trained with cross-validation saw the smallest gap in RMSE, indicating this model strongly generalises on unseen data.
- Unpruned Tree model had the largest gap between train and test RMSE, indicating this model is more prone to overfitting on unseen data.
- Both pruned and unpruned tree saw the same RMSE on training data, however the pruned tree was less error-prone on the test data, showing the effectiveness of pruning the decision tree. 

### Pre-pruning Exploration
```{r prepruning exploration, echo=FALSE, warning=FALSE}
preprune_config <- list(list(maxdepth = 3, minsplit = 10, minbucket = 5),
                        list(maxdepth = 5, minsplit = 10, minbucket = 5),
                        list(maxdepth = 10, minsplit = 20, minbucket = 10),
                        list(maxdepth = 15, minsplit = 30, minbucket = 15))

prepruning_results <- lapply(preprune_config, function(configs) {
  tree_preprune <- rpart(
    life.quality ~ .,
    data = train[, c(predictors, "life.quality")],
    method = "anova",
    control = rpart.control(maxdepth = configs$maxdepth, minsplit = configs$minsplit, minbucket = configs$minbucket, cp = 0.01)
  )
  pred_preprune <- predict(tree_preprune, test[, predictors])
  data.frame(
    MaxDepth = configs$maxdepth, 
    MinSplit = configs$minsplit, 
    MinBucket = configs$minbucket,
    RMSE = round(rmse(test$life.quality, pred_preprune), 4),
    R2 = round(r2(test$life.quality, pred_preprune), 4),
    N_leaves = sum(tree_preprune$frame$var == "<leaf>")
  )
})

preprune_df <- do.call(rbind, prepruning_results)
preprune_df <- preprune_df[order(preprune_df$RMSE), ]
print(preprune_df)
```
#### Discussion
- Pre-pruning reveals that the decision tree naturally stops growth, produces 7-8 leaves regardless of what is specified as the maximum depth of the tree. 
- Consistently low $R^2$ value, ranging from ~0.03 to ~0.05. Ties back into the point raised about ceiling effects
- All variants reveal similar RMSE values

### Residual Analysis
```{r residual plots, echo=FALSE, warning=FALSE}
prune_residuals <- test$life.quality - pred_prun

par(mfrow = c(1,2))

plot(pred_prun, prune_residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residuals vs Fitted (Pruned Tree)")
abline(h = 0, col = "red", lty = 2, lwd = 2)

hist(prune_residuals, breaks = 30, main = "Residual Distribution (Pruned Tree)", xlab = "Residual")
abline(v = 0, col = "red", lty = 2, lwd = 2)
```

```{r qqplot, echo=FALSE, warning=FALSE}
qqnorm(prune_residuals, main = "QQ Plot of Residuals (Pruned Tree)", pch = 20)
qqline(prune_residuals, col = "red", lwd = 2)
```

```{r residual summary stats, echo=FALSE, warning=FALSE}
cat("Residual Summary:\n")
print(summary(prune_residuals))
cat(sprintf("\nMean Residual: %.4f\n", mean(prune_residuals)))
cat(sprintf("SD of Residuals: %.4f", sd(prune_residuals)))
```
#### Discussion
- Residual vs fitted plot displays no systemic bias, points follow shape of vertical columns but this is expected from survey data and decision tree nature (fixed leaf nodes).
- Residual distribution shows a very close bell-shape, mostly centered near zero. 
- QQ Plot confirms near-normality of data, points follow diagonal very closely in the middle and minimal deviation at both tails of the graph
- Mean residual of 0.098 shows slight tendency to under-predict QoL.
- Plots show that model is overall valid, especially with how distribution of residuals how the QQ Plot is shaped. 

### Test-Retest Reliability
```{r test-retest reliability, echo=FALSE, warning=FALSE}
set.seed(123)
train_idx_retest <- sample(nrow(sub_data), 0.8 * nrow(sub_data))
train_retest <- sub_data[train_idx_retest, ]
test_retest  <- sub_data[-train_idx_retest, ]

tree_pruned_retest <- rpart(
  life.quality ~ .,
  data      = train_retest[, c(predictors, "life.quality")],
  method    = "anova",       # use "class" for classification
  control   = rpart.control(cp = 0.01)
)

imp_rank <- tree_pruned$variable.importance
imp_rank_retest <- tree_pruned_retest$variable.importance

vars_combined <- union(names(imp_rank), names(imp_rank_retest))
imp_rank_aligned <- imp_rank[vars_combined]
imp_rank_retest_aligned <- imp_rank_retest[vars_combined]

# Remove NA values
imp_rank_aligned[is.na(imp_rank_aligned)] <- 0
imp_rank_retest_aligned[is.na(imp_rank_retest_aligned)] <- 0

# Normalise data
imp_rank_norm <- 100 * imp_rank_aligned / max(imp_rank_aligned)
imp_rank_retest_norm <- 100 * imp_rank_retest_aligned / max(imp_rank_retest_aligned)

test_retest_df <- data.frame(Variable = vars_combined, Split1 = round(imp_rank_norm, 2), Split2 = round(imp_rank_retest_norm, 2)) |> 
  arrange(desc(Split1))

print(test_retest_df)
```

```{r importance rank correlation, echo=FALSE, warning=FALSE}
imp_rank_cor <- cor(rank(-imp_rank_norm), rank(-imp_rank_retest_norm), method="spearman")
cat(sprintf("Feature Importance Rank Correlation (Spearman): %.3f\n", imp_rank_cor))
```

```{r rank visual, echo=FALSE, warning=FALSE}
test_retest_df_long <- test_retest_df |>
  pivot_longer(cols = c("Split1", "Split2"), names_to = "Split", values_to = "Importance")
ggplot(test_retest_df_long, aes(x = reorder(Variable, Importance), y = Importance, fill = Split)) + 
  geom_bar(stat="identity", position="dodge") + labs(title = "Test-Retest: Feature Importance between two Different Data Splits", x = "Variable", y = "Normalised Importance")
```
#### Discussion
- Strong consistency in rankings between the two splits
- `gone_no_health` the most dominant predictor (scoring 100 in both splits)
- `gone_no_income` and `gone_no_food` demonstrated strong stability in both splits, minimal score deviation from first split to second split (~1 point difference for `gone_no_income` and ~2 point difference for `gone_no_food`)
- `trust_neighbourhood` and `social_class` showed greater variation in feature importance score, still retained same rank in both splits.

## Limitations

## References

