---
title: "Group1"
output:
  pdf_document: default
  html_document: default
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 5)
```

## Introduction

## Literature Review

## Theory

### Hypothesis

## Exploratory Data Analysis
```{r echo=FALSE, warning=FALSE}
library(tidyverse)
library(psych)
library(ggpubr)

data <- read.csv("Data/wvs_data.csv")
```

```{r echo=FALSE, warning=FALSE}
data <- data %>%
  dplyr::mutate(happiness = dplyr::recode(V10, 
                                   '1' = 4,
                                   '2' = 3,
                                   '3' = 2,
                                   '4' = 1, 
                                   '-2' = NULL),
         
         health = dplyr::recode(V11,
                                '1' = 5, 
                                '2' = 4, 
                                '3' = 3,
                                '4' = 2,
                                '5' = 1, 
                                '-2' = NULL),
         
         finances = dplyr::recode(V59, 
                                  '-2' = NULL,
                                  .default = V59),
         
         satisfaction = dplyr::recode(V23, 
                                  '-2' = NULL,
                                  .default = V23),
         
         freedom = dplyr::recode(V55, 
                                  '-2' = NULL,
                                  .default = V55),
         gone_no_food = dplyr::recode(V188,
                                      '-2' = NULL,
                                      .default = V188),
         gone_no_health = dplyr::recode(V190,
                                        '-2' = NULL,
                                        .default = V190),
         gone_no_income = dplyr::recode(V191,
                                        '-2' = NULL,
                                        .default = V191),
         trust_neighborhood = dplyr::recode(V103,
                                            '-2' = NULL,
                                            .default = V103),
         social_class = dplyr::recode(V236,
                                      '-2' = NULL,
                                      .default = V236))

fa.fit <- fa(data[,c("happiness",
                         "health",
                         "finances", 
                         "satisfaction",
                         "freedom")], 
             nfactors=1)

data$life.quality <- as.numeric(fa.fit$scores)

cols <- c("gone_no_food", "gone_no_health", "gone_no_income", "trust_neighborhood", "social_class", "life.quality")
sub_data <- data %>% select(all_of(cols))
```

### Missing Data Plot
```{r echo=FALSE, warning=FALSE}
na_counts <- colSums(is.na(sub_data[, cols]))

barplot(na_counts,
        main = "Missing Values per Column",
        xlab = "Column",
        ylab = "Count of NAs",
        col = "steelblue",
        las = 2)  # rotate x labels
```

### Data Cleaning
```{r echo=FALSE, warning=FALSE}
sub_data <- sub_data[!is.na(sub_data$life.quality), ]
predictors <- setdiff(cols, "life.quality")

sub_data <- na.omit(sub_data)
sub_data$life.quality <- as.numeric(sub_data$life.quality)
sub_data[, predictors] <- lapply(sub_data[, predictors], as.numeric)
```

## Methodology

### Assumptions

### Testing and Validation
#### Unpruned and Pruned trees comparison
```{r echo=FALSE, warning=FALSE}
library(rpart)
library(rpart.plot)

train_idx <- sample(nrow(sub_data), 0.8 * nrow(sub_data))
train <- sub_data[train_idx, ]
test  <- sub_data[-train_idx, ]

tree_unpruned <- rpart(
  life.quality ~ .,
  data      = train[, c(predictors, "life.quality")],
  method    = "anova",       # use "class" for classification
  control   = rpart.control(cp = 0, minsplit = 2, minbucket = 1)
)
best_cp <- tree_unpruned$cptable[
  which.min(tree_unpruned$cptable[, "xerror"]), "CP"
]

tree_pruned <- prune(tree_unpruned, cp = best_cp)
# ── 4. Compare on test set
pred_unp  <- predict(tree_unpruned, test[, predictors])
pred_prun <- predict(tree_pruned,   test[, predictors])
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
```

#### K-Folds cross validation
```{r echo=FALSE, warning=FALSE}
library(caret)

set.seed(123)
# ── 1. Set up k-fold CV (k=10 is standard)
ctrl <- trainControl(method = "cv", number = 10)

# ── 2. Train with CV
cv_tree <- train(
  life.quality ~ .,
  data      = train[, c(predictors, "life.quality")],
  method    = "rpart",
  trControl = ctrl,
  tuneLength = 10        
)

plot(cv_tree)            

pred_cv <- predict(cv_tree, newdata = test[, predictors])

```

#### Bagging validation:
```{r echo=FALSE, warning=FALSE}
library(rpart)

set.seed(123)
n_bags  <- 10          # number of bootstrap samples
n_train <- nrow(train)

# Store each tree
trees <- vector("list", n_bags)

for (i in 1:n_bags) {
  # Bootstrap sample
  idx       <- sample(1:n_train, n_train, replace = TRUE)
  boot_data <- train[idx, c(predictors, "life.quality")]
  
  # Fit tree on bootstrap sample
  trees[[i]] <- rpart(
    life.quality ~ .,
    data    = boot_data,
    method  = "anova",
    control = rpart.control(cp = 0.01)
  )
}

preds_matrix <- sapply(trees, function(tree) {
  predict(tree, newdata = test[, predictors])
})

# Average predictions across all trees (this is bagging)
pred_bag <- rowMeans(preds_matrix)

par(mfrow = c(2, 2))   # show 6 trees
for (i in 1:2) {
  rpart.plot(trees[[i]],
             main = paste("Bootstrap Tree", i),
             cex  = 0.6,
             type = 2)
}
par(mfrow = c(1, 1))
```
Where we try to observe the different decision trees formed by using bootstrap samples of the dataset provided for our interested features.
## Result Analysis

### Feature Importance observed through K-folds validation where the number of folds = 10
```{r echo=FALSE, warning=FALSE}
plot(varImp(cv_tree),
     main = "Feature Importance - 10-Fold cross validation Tree",
     top  = 10)  
```

### The goodness of fit ($R^{2}$) for each model and the root mean squared error for each model
```{r echo=FALSE, warning=FALSE}
r2 <- function(actual, pred) {
  ss_res <- sum((actual - pred)^2)
  ss_tot <- sum((actual - mean(actual))^2)
  1 - (ss_res / ss_tot)
}

results <- data.frame(
  Model = c("Unpruned Tree", "Pruned Tree", "CV Tree", "Bagging"),
  RMSE  = round(c(
    rmse(test$life.quality, pred_unp),
    rmse(test$life.quality, pred_prun),
    rmse(test$life.quality, pred_cv),
    rmse(test$life.quality, pred_bag)
  ), 4),
  R2 = round(c(
    r2(test$life.quality, pred_unp),
    r2(test$life.quality, pred_prun),
    r2(test$life.quality, pred_cv),
    r2(test$life.quality, pred_bag)
  ), 4)
)

# Sort by RMSE ascending (best first)
results <- results[order(results$RMSE), ]

print(results)
```

### Plotting goodness of fit for each feature $R^{2}$ used in the decision tree.
```{r echo=FALSE, warning=FALSE}
# R² per feature = correlation² between feature and life.quality
feature_r2 <- sapply(predictors, function(var) {
  cor(as.numeric(test[[var]]), test$life.quality, use = "complete.obs")^2
})

# Convert to data frame
feature_r2_df <- data.frame(
  Variable = names(feature_r2),
  R2       = round(feature_r2, 4)
) |> arrange(desc(R2))

ggplot(feature_r2_df, aes(x = reorder(Variable, R2), y = R2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "R² per Feature vs Life Quality",
       x     = "Variable",
       y     = "R²") +
  theme_minimal()
```

## Limitations

## References

