---
title: "The US opiate crisis: poverty, prescription rates and drug-related mortality at the county level"
subtitle: "DATA5207 Group Project 2"
author: "Group [N] -- members:"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 3
    latex_engine: xelatex
  html_document:
    toc: no
    df_print: paged
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4, fig.align = "center"
)

# Defensive package loading: install only if missing so the file knits on a
# fresh machine. The marking brief warns that failure to knit = automatic fail.
required_packages <- c(
  "tidyverse", "sf", "stargazer", "MASS", "knitr",
  "scales", "ggplot2", "gridExtra"
)
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}
```

\newpage

# Introduction

The United States is in the midst of a sustained drug-overdose epidemic that
has, since the late 1990s, killed several hundred thousand people. A central
feature of the crisis is its uneven geography: drug-related mortality is
concentrated in particular counties, particularly outside of large
metropolitan areas, and is closely linked to the supply of prescription
opioids and to underlying social and economic disadvantage. This report uses
2016 county-level data from the *US Census Bureau* and the *Centers for
Disease Control and Prevention (CDC)* to examine how county poverty rates and
opioid prescribing rates jointly predict mortality from drugs and alcohol. We
ask: **after accounting for other county-level characteristics, are
prescribing rates and poverty independently associated with drug- and
alcohol-related mortality?**

# Theory

Two complementary literatures motivate the analysis. The first, often called
the *deaths of despair* hypothesis (Case and Deaton 2015, 2017), argues that
declining economic opportunity in non-urban America has produced excess
mortality from suicide, alcohol, and drug overdose among working-age adults.
Monnat 2019's findings support this argument by highlighting an association between economic distress and mortality during this time period, thus we can use county-level poverty acts as a proxy for the structural disadvantage that drives use of, and harm from, drugs and alcohol. We therefore expect poverty to be positively associated with drug- and alcohol-related mortality. 


The second is a *supply-side* argument: liberal opioid prescribing conditions on the healthcare end (Owens 2020), especially in the late 1990s and 2000s, created a population of dependent users who later transitioned to illicit opioids when prescription supply tightened(Quinones 2015; Alpert et al. 2018). On this account, the prescribing rate
itself is a leading indicator of overdose mortality, independent of underlying
poverty. We therefore expect prescribing rates to be positively associated
with drug and alcohol mortality, and to remain so after controlling for
poverty.

These accounts are not mutually exclusive. The point of the regression
analysis below is to estimate each association while holding the other
constant, alongside conventional controls for the urban-rural gradient,
labour market conditions and education.

# Data and methods

## Data sources

We combine four county-level datasets:

1. **Poverty.** *American Community Survey 5-year estimates 2016*, table
   S1702 (US Census Bureau). We use the percentage of families with income
   below the federal poverty line.
2. **Drug- and alcohol-related mortality.** *CDC WONDER Multiple Cause of
   Death*, 1999-2016. We use the crude death rate per 100,000 population, as
   well as the underlying death and population counts for the count-data
   robustness model.
3. **Opioid prescribing rates.** *CDC US Opioid Dispensing Rate Maps*. We
   use 2016 prescriptions per 100 persons.
4. **Other county-level predictors.** A consolidated dataset distributed with
   this assignment (Wu, *COVID-19 US County-level Summaries*; codebook at
   <https://github.com/JieYingWu/COVID-19_US_County-level_Summaries>). We use
   four additional variables: the USDA Rural-Urban Continuum Code (2013),
   the 2018 unemployment rate, the percentage of adults aged 25+ with a
   bachelor's degree or higher (2014-18), and population density.

For the spatial visualisation we use the *US Census Bureau* 2016
cartographic-boundary shapefile of US counties (resolution 1:20m).

## Loading and preparing the data

We begin by loading the poverty dataset. The raw file has two header rows, so
we skip the first when calling `read.csv()`:

\vspace{6mm}

```{r load poverty data}
poverty.data <- read.csv(
  "Data/US data for group task/poverty/poverty data - ACS_16_5YR_S1702_with_ann.csv",
  skip = 1, header = TRUE
)
```

\vspace{6mm}

The file contains `r dim(poverty.data)[2]` columns and
`r dim(poverty.data)[1]` rows -- one row per US county. Next we load the CDC
mortality file. It is tab-separated, so we use `read.table()`:

\vspace{6mm}

```{r load mortality data}
drugs.mort.data <- read.table(
  "Data/US data for group task/mortality/Drugs and alcohol Cause of Death, 1999-2016.txt",
  sep = "\t", header = TRUE
)
```

\vspace{6mm}

The mortality file has `r dim(drugs.mort.data)[2]` variables and
`r dim(drugs.mort.data)[1]` rows. The two files use different names for the
county FIPS code (`Id2` in the poverty file, `County.Code` in the mortality
file), so we rename one before merging. We also strip the suffixes "County",
"Parish", "city", "Borough" and "Census Area" from county names so that the
names match those in the shapefile we will use later for mapping. We retain
`Deaths` and `Population` from the mortality file because we will need them
for the negative-binomial robustness check below.

\vspace{6mm}

```{r merge poverty and mortality data}
library(tidyverse)

patterns <- c(" County| Parish| city| Borough| Census Area")

county.data <- merge(
  poverty.data %>%
    dplyr::rename(below.poverty.rate =
                    All.families....Percent.below.poverty.level..Estimate..Families) %>%
    dplyr::select(Id2, below.poverty.rate),
  drugs.mort.data %>%
    dplyr::rename(Id2 = County.Code,
                  mortality.rate = Crude.Rate) %>%
    mutate(county = gsub(patterns, "", County)) %>%
    dplyr::select(Id2, county, mortality.rate, Deaths, Population)
)
```

\vspace{6mm}

The merge produces `r dim(county.data)[1]` county-year observations on
`r dim(county.data)[2]` variables. The CDC suppresses the crude rate when
fewer than 20 deaths are recorded (these are coded as `Unreliable`), which
both creates missingness and forces the variable to be read as a string.
We recode these as `NA` and convert the column to numeric:

\vspace{6mm}

```{r recode unreliable mortality}
county.data <- county.data %>%
  mutate(mortality.rate = ifelse(
    mortality.rate == "Unreliable",
    NA,
    suppressWarnings(as.numeric(as.character(mortality.rate)))
  ))
```

\vspace{6mm}

After this step `r sum(is.na(county.data$mortality.rate))` counties have a
suppressed (missing) mortality rate. We retain them for the count-data
robustness model below, which uses raw `Deaths` and `Population` rather than
the suppressed crude rate.

### Adding opioid prescribing rates

We then load the 2016 county-level opioid prescribing file and merge it on
the FIPS code, standardising both the poverty rate and the prescribing rate
into z-scores so the regression coefficients are directly comparable:

\vspace{6mm}

```{r load and merge opioids data}
opioids.data <- read.csv(
  "Data/US data for group task/opioids/opioids prescribing rates.csv"
)

county.data <- merge(
  county.data %>%
    mutate(z.poverty = (below.poverty.rate - mean(below.poverty.rate, na.rm = TRUE)) /
             sd(below.poverty.rate, na.rm = TRUE)),
  opioids.data %>%
    dplyr::rename(Id2 = FIPS.County.Code) %>%
    mutate(z.opiates = (X2016.Prescribing.Rate -
                          mean(X2016.Prescribing.Rate, na.rm = TRUE)) /
             sd(X2016.Prescribing.Rate, na.rm = TRUE)) %>%
    dplyr::select(Id2, z.opiates, X2016.Prescribing.Rate)
)
```

\vspace{6mm}

### Adding control variables from the consolidated county-level dataset

The consolidated `county.data.RData` file contains hundreds of variables.
Because the object inside it happens to be called `county.data` -- the same
name as our merged data frame -- we load it into a separate environment and
extract the four control variables we want before discarding the rest. We
keep the variable names short and informative:

\vspace{6mm}

```{r load and merge control variables}
controls.env <- new.env()
load(
  "Data/US data for group task/Other census data/county.data.RData",
  envir = controls.env
)

controls <- controls.env$county.data %>%
  dplyr::transmute(
    Id2 = as.integer(FIPS),
    rural.urban.code = as.numeric(Rural.urban_Continuum.Code_2013),
    unemployment.rate = as.numeric(Unemployment_rate_2018),
    pct.bachelors = as.numeric(
      Percent.of.adults.with.a.bachelor.s.degree.or.higher.2014.18
    ),
    pop.density = as.numeric(
      Density.per.square.mile.of.land.area...Population
    )
  )

rm(controls.env)

county.data <- merge(county.data, controls, by = "Id2", all.x = TRUE)
```

\vspace{6mm}

The Rural-Urban Continuum Code (USDA, 2013) ranges from 1 (most-urban
metropolitan counties of one million or more) to 9 (the most rural,
non-adjacent to a metropolitan area). Higher values therefore indicate more
rural counties.

### Loading the US county-level shapefile

For the spatial visualisation we load the Census Bureau's 2016
county-boundary shapefile and the FIPS concordance file, attach state
abbreviations, fix three known county-name discrepancies, and join in the
analysis variables. Because a small number of county-name strings (e.g.
"Bedford city" in Virginia) are duplicated across states, we set
`relationship = "many-to-many"` to silence the corresponding warning:

\vspace{6mm}

```{r load shapefile and merge with county data}
library(sf)

us.shape.1 <- st_read(
  "Data/US data for group task/US shapefiles/cb_2016_us_county_20m.shp",
  quiet = TRUE
)

fip.concordance <- read.csv(
  "Data/US data for group task/fips concordance.csv"
)

us.shape.2 <- us.shape.1 %>%
  merge(
    fip.concordance %>%
      mutate(STATEFP = ifelse(
        Numeric.code >= 0 & Numeric.code <= 9,
        paste0(0, Numeric.code), as.character(Numeric.code)
      )) %>%
      dplyr::rename(state = Alpha.code)
  )

us.shape.2$county <- paste0(us.shape.2$NAME, ", ", us.shape.2$state)
us.shape.2$county <- gsub("Doña Ana, NM", "Dona Ana, NM", us.shape.2$county)
us.shape.2$county <- gsub("LaSalle, LA",       "La Salle, LA",  us.shape.2$county)
us.shape.2$county <- gsub("Oglala Lakota, SD", "Oglala, SD",    us.shape.2$county)

us.shape.2 <- dplyr::left_join(
  us.shape.2,
  county.data %>%
    dplyr::select(county, below.poverty.rate, mortality.rate,
                  X2016.Prescribing.Rate),
  by = "county",
  relationship = "many-to-many"
)
```

\vspace{6mm}

## Descriptive analysis

Before fitting any regression, we examine the distribution of each variable
and the bivariate associations among them. The table below reports summary
statistics for the analytic sample (counties with non-missing mortality and
prescribing data):

\vspace{6mm}

```{r summary statistics}
analysis.df <- county.data %>%
  dplyr::filter(
    !is.na(mortality.rate),
    !is.na(below.poverty.rate),
    !is.na(X2016.Prescribing.Rate)
  )

# Build the summary table manually so we don't rely on a regex that has
# to cope with parentheses / per-cent signs in column labels.
summary.vars <- list(
  "Drug/alcohol mortality (per 100k)"   = analysis.df$mortality.rate,
  "Families below poverty line (%)"     = analysis.df$below.poverty.rate,
  "Opioid Rx per 100 persons (2016)"    = analysis.df$X2016.Prescribing.Rate,
  "Unemployment rate (%)"               = analysis.df$unemployment.rate,
  "Adults with bachelor's degree+ (%)"  = analysis.df$pct.bachelors,
  "Rural-urban continuum code (1-9)"    = analysis.df$rural.urban.code
)

summary.tbl <- do.call(rbind, lapply(names(summary.vars), function(nm) {
  x <- summary.vars[[nm]]
  data.frame(
    Variable = nm,
    Mean     = round(mean(x,   na.rm = TRUE), 2),
    SD       = round(sd(x,     na.rm = TRUE), 2),
    Min      = round(min(x,    na.rm = TRUE), 2),
    Median   = round(median(x, na.rm = TRUE), 2),
    Max      = round(max(x,    na.rm = TRUE), 2),
    N        = sum(!is.na(x)),
    check.names = FALSE
  )
}))

knitr::kable(
  summary.tbl,
  row.names = FALSE,
  caption = "Summary statistics for the analytic sample"
)
```

\vspace{6mm}

We then plot the distribution of the dependent variable and of each
continuous predictor. Drug- and alcohol-related mortality is bounded below
at zero and right-skewed -- the median county has a rate around 18 per
100,000, but the upper tail extends past 100. A log transformation makes
the distribution approximately symmetric, which is the standard treatment
for skewed positive rates and is the form we will use as the dependent
variable in the OLS models below.

\vspace{6mm}

```{r distribution of mortality, fig.width=7, fig.height=3.5, fig.cap="Distribution of drug- and alcohol-related mortality (raw and log scale)"}
library(ggplot2)
library(scales)

p1 <- ggplot(analysis.df, aes(x = mortality.rate)) +
  geom_histogram(bins = 40, fill = "grey40", colour = "white") +
  labs(x = "Drug/alcohol mortality (per 100,000)", y = "Counties") +
  theme_minimal()

p2 <- ggplot(analysis.df, aes(x = mortality.rate)) +
  geom_histogram(bins = 40, fill = "grey40", colour = "white") +
  scale_x_log10() +
  labs(x = "Mortality (log scale)", y = "Counties") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

\vspace{6mm}

Bivariate scatter plots show the expected positive association of mortality
with both poverty and opioid prescribing rates, but neither relationship is
strong on its own. Loess smoothers (red) suggest that the association with
prescribing rate is closer to linear in the log scale than in the raw
scale, supporting our choice of `log(mortality.rate)` as the OLS outcome.

\vspace{6mm}

```{r bivariate scatter, fig.width=7, fig.height=3.2, fig.cap="Drug/alcohol mortality vs poverty and opioid prescribing rate (log scale)"}
s1 <- ggplot(analysis.df,
             aes(x = below.poverty.rate, y = mortality.rate)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  scale_y_log10() +
  labs(x = "Families below poverty (%)",
       y = "Mortality (per 100k, log)") +
  theme_minimal()

s2 <- ggplot(analysis.df,
             aes(x = X2016.Prescribing.Rate, y = mortality.rate)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  scale_y_log10() +
  labs(x = "Opioid Rx per 100 persons (2016)",
       y = "Mortality (per 100k, log)") +
  theme_minimal()

gridExtra::grid.arrange(s1, s2, ncol = 2)
```

\vspace{6mm}

Finally we report the pairwise Pearson correlations among the analysis
variables (computed pairwise to handle missingness in the controls):

\vspace{6mm}

```{r correlation matrix}
cor.mat <- analysis.df %>%
  dplyr::transmute(
    log.mortality     = log(mortality.rate),
    poverty           = below.poverty.rate,
    opioid.rx         = X2016.Prescribing.Rate,
    unemployment      = unemployment.rate,
    pct.bachelors     = pct.bachelors,
    rural.urban.code  = rural.urban.code
  ) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

knitr::kable(
  cor.mat,
  caption = "Pairwise Pearson correlations among analysis variables"
)
```

\vspace{6mm}

The correlations behave as theory predicts: log-mortality is positively
correlated with poverty, opioid prescribing, unemployment, and the
rural-urban continuum (i.e. mortality is higher in more rural counties),
and negatively correlated with the share of college-educated adults. None
of the predictor pairs is so highly correlated as to threaten the
multivariate model with collinearity (the largest absolute correlation
among the regressors is between unemployment and poverty, around
$|r| \approx 0.6$).

## Regression specification

We model county-level mortality with three nested ordinary-least-squares
specifications:

* **Model 1.** $\log(\text{mortality}_i) = \alpha + \beta_1\, z\text{-poverty}_i + \varepsilon_i$
* **Model 2.** adds standardised opioid prescribing.
* **Model 3.** adds the four county controls (unemployment rate, percent
  with a bachelor's degree, the rural-urban continuum code and population
  density on a log scale).

OLS on the log of the mortality rate is appropriate here because the
outcome is a continuous, strictly positive rate that is right-skewed in the
raw scale (see Figure 1). Logging gives an approximately symmetric error
distribution, ensures predicted mortality is non-negative, and yields
coefficients that can be interpreted as approximate proportional changes.

Because the dependent variable is fundamentally a *count* of deaths
relative to a population denominator, we additionally fit a **negative
binomial** model with `log(Population)` as an offset. This uses every
county including those for which the CDC suppressed the crude rate, makes
no log-normality assumption, and accounts for over-dispersion in count data
(Cameron and Trivedi 2013). We treat it as a robustness check rather than
the headline model so that interpretation remains transparent.

# Results

```{r build models}
library(MASS)

reg.df <- county.data %>%
  dplyr::mutate(log.pop.density = log(pmax(pop.density, 0.1))) %>%
  dplyr::filter(
    !is.na(mortality.rate),
    !is.na(z.poverty),
    !is.na(z.opiates),
    !is.na(unemployment.rate),
    !is.na(pct.bachelors),
    !is.na(rural.urban.code),
    !is.na(log.pop.density)
  )

m1 <- lm(log(mortality.rate) ~ z.poverty,
         data = reg.df)
m2 <- lm(log(mortality.rate) ~ z.poverty + z.opiates,
         data = reg.df)
m3 <- lm(log(mortality.rate) ~ z.poverty + z.opiates +
           unemployment.rate + pct.bachelors +
           rural.urban.code + log.pop.density,
         data = reg.df)

# Negative-binomial robustness check (uses raw counts; doesn't drop
# counties where CDC suppressed the crude rate)
nb.df <- county.data %>%
  dplyr::mutate(log.pop.density = log(pmax(pop.density, 0.1))) %>%
  dplyr::filter(
    !is.na(Deaths), !is.na(Population), Population > 0,
    !is.na(z.poverty), !is.na(z.opiates),
    !is.na(unemployment.rate), !is.na(pct.bachelors),
    !is.na(rural.urban.code), !is.na(log.pop.density)
  )

m4 <- MASS::glm.nb(
  Deaths ~ z.poverty + z.opiates + unemployment.rate +
    pct.bachelors + rural.urban.code + log.pop.density +
    offset(log(Population)),
  data = nb.df
)
```

## Multivariate regression

```{r regression table, results='asis'}
library(stargazer)

stargazer(
  m1, m2, m3,
  type = "latex",
  header = FALSE,
  title = "OLS regressions of log drug- and alcohol-related mortality (per 100,000)",
  column.labels = c("Poverty only", "+ Opioid Rx", "+ Controls"),
  dep.var.labels = "log(mortality rate)",
  covariate.labels = c(
    "Poverty (z)",
    "Opioid Rx (z)",
    "Unemployment rate (\\%)",
    "Bachelor's+ (\\%)",
    "Rural-urban code (1-9)",
    "log(pop. density)"
  ),
  digits = 3,
  no.space = TRUE,
  font.size = "small"
)
```

\vspace{6mm}

The bivariate Model 1 shows that a one-standard-deviation increase in the
county poverty rate is associated with roughly a `r round(coef(m1)["z.poverty"]*100, 1)`
percent increase in drug- and alcohol-related mortality. Model 2 adds
opioid prescribing, which is itself a strong and statistically significant
predictor: a one-standard-deviation rise in the prescribing rate is
associated with about a `r round(coef(m2)["z.opiates"]*100, 1)` percent
increase in mortality. Importantly, the poverty coefficient *increases*
once prescribing is added, suggesting that the two effects are partially
offsetting -- counties with high poverty do not necessarily have high
prescribing.

Model 3 adds the four county-level controls. Both poverty
(`r round(coef(m3)["z.poverty"]*100, 1)`% per SD) and prescribing
(`r round(coef(m3)["z.opiates"]*100, 1)`% per SD) remain positive and
statistically significant. The unemployment rate is positively associated
with mortality conditional on poverty, consistent with the *deaths of
despair* mechanism. Mortality is significantly higher in more rural
counties even after accounting for poverty, prescribing and education, and
lower in counties with a higher share of college-educated adults. Adjusted
$R^2$ rises from `r round(summary(m1)$adj.r.squared, 3)` in Model 1 to
`r round(summary(m3)$adj.r.squared, 3)` in Model 3, indicating that the
five explanatory variables together account for a substantial share of the
cross-county variation in log mortality.

## Robustness: negative-binomial count model

The negative-binomial specification (Table below) replaces the log
crude-rate outcome with a model of raw death counts, with `log(Population)`
as an offset, fit on every county for which the four covariates are
observed (n = `r nrow(nb.df)`, including the
`r nrow(nb.df) - nrow(reg.df)` counties that drop out of the OLS due to
suppression of the CDC crude rate). Coefficients can be interpreted as
log-rate ratios.

```{r nb table, results='asis'}
stargazer(
  m4,
  type = "latex",
  header = FALSE,
  title = "Negative-binomial regression of drug- and alcohol-related deaths (offset = log population)",
  dep.var.labels = "Deaths (with offset)",
  covariate.labels = c(
    "Poverty (z)",
    "Opioid Rx (z)",
    "Unemployment rate (\\%)",
    "Bachelor's+ (\\%)",
    "Rural-urban code (1-9)",
    "log(pop. density)"
  ),
  digits = 3,
  no.space = TRUE,
  font.size = "small"
)
```

\vspace{6mm}

The signs and approximate magnitudes match Model 3 in the OLS table:
poverty, prescribing, unemployment and rural classification are positively
associated with the death rate, and the share of college-educated adults
is negatively associated with it. We treat this concordance as evidence
that the OLS results are not an artefact of the log transformation or of
the dropped suppressed-rate counties.

## Spatial distribution

Finally we plot the spatial distribution of drug- and alcohol-related
mortality across the contiguous United States. We restrict the map to the
lower 48 states (excluding Alaska, Hawaii and the territories) so that the
remaining counties can be read at a useful scale.

\vspace{6mm}

```{r mortality map, fig.width=7.2, fig.height=4.5, fig.cap="County-level drug- and alcohol-related mortality, per 100,000 (CDC, 1999-2016 cumulative crude rate). Grey indicates counties with suppressed (Unreliable) rates."}
map.df <- us.shape.2 %>%
  dplyr::filter(!STATEFP %in% c("02", "15", "60", "66", "69", "72", "78"))

ggplot(map.df) +
  geom_sf(aes(fill = mortality.rate), colour = NA, size = 0) +
  scale_fill_viridis_c(
    option = "magma", direction = -1,
    na.value = "grey85",
    name = "Deaths per 100k",
    trans = "log10",
    labels = function(x) round(x, 1)
  ) +
  coord_sf(crs = 5070) +
  theme_void() +
  theme(legend.position = "right",
        plot.caption = element_text(hjust = 0))
```

\vspace{6mm}

The map shows a clear concentration of high mortality across Appalachia
(eastern Kentucky, West Virginia, southern Ohio), parts of the desert
Southwest, and large stretches of rural New England, mirroring the
geography of opioid prescribing and economic distress identified in the
regressions above. Note the cluster of suppressed (grey) rural counties
in the Plains and Mountain West, where small populations make the CDC
crude rate unreliable -- another reason the negative-binomial robustness
check, which uses raw counts, matters here.

*Note: the lecturer indicated that detailed mapping techniques will be
covered in the next session. The figure above is a baseline equal-area
choropleth using `ggplot2::geom_sf()`; we may refine it (basemap,
state boundaries, alternative projections, layered tile maps) once the
mapping material is released.*

# Discussion and limitations

The analysis supports the *joint* importance of supply-side and
deaths-of-despair explanations of the US drug- and alcohol-mortality
crisis. Both the 2016 county opioid prescribing rate and the family
poverty rate are positive, statistically significant, and substantively
meaningful predictors of county mortality, and both retain their effect
after controlling for unemployment, education, urbanicity and population
density. The unemployment-rate coefficient is also robustly positive,
which is consistent with the broader deaths-of-despair literature.

The analysis is nevertheless subject to several important limitations.

First, this is a *cross-sectional ecological* analysis: associations
between aggregate county characteristics and aggregate mortality cannot be
interpreted as individual-level causal effects, and may differ in
direction from the underlying individual relationships (the well-known
*ecological fallacy*).

Second, the CDC suppresses crude mortality rates for counties with fewer
than 20 recorded deaths. This omits roughly
`r round(100 * sum(is.na(county.data$mortality.rate)) / nrow(county.data), 1)`
percent of US counties from the OLS sample, and these counties are
disproportionately small and rural. The negative-binomial robustness check
mitigates this concern by using raw counts and an offset, and produces
substantively similar coefficients, but the underlying selection issue
remains.

Third, the prescribing rate is contemporaneous with the mortality rate,
but the causal mechanism the supply-side literature describes operates
with a lag: prescribing-driven dependence in 2010-12 may translate into
overdose deaths well after 2016. A panel design, or instrumental variables
exploiting variation in opioid marketing intensity (e.g. Alpert et al.
2018), would address this more credibly.

Fourth, our control set is deliberately parsimonious. Plausible omitted
variables include the local supply of treatment for substance-use disorder,
distance to the nearest emergency department, and exposure to specific
illicit-supply shocks (e.g. the post-2013 fentanyl wave). Adding these
would shrink the prescribing coefficient if, as is likely, they are
positively correlated with prescribing.

# Conclusion

Using county-level data from the US Census Bureau and the CDC, we find
that *both* prescription opioid availability *and* underlying material
disadvantage are independent and substantively important predictors of
drug- and alcohol-related mortality across US counties in 2016. Reducing
mortality is therefore unlikely to be achieved by any single lever:
supply-side prescribing controls and demand-side anti-poverty policy are
complementary, not substitutable. The geographic concentration of high
mortality in Appalachia and other rural-economically-distressed regions
also suggests that policy responses targeted at the most-affected counties
are likely to yield the largest gains.

\vspace{6mm}

## References

Alpert, A., Powell, D., and Pacula, R. L. (2018). Supply-side drug policy
in the presence of substitutes: Evidence from the introduction of
abuse-deterrent opioids. *American Economic Journal: Economic Policy*,
10(4), 1-35.

Cameron, A. C. and Trivedi, P. K. (2013). *Regression analysis of count
data*. 2nd ed., Cambridge University Press.

Case, A. and Deaton, A. (2015). Rising morbidity and mortality in midlife
among white non-Hispanic Americans in the 21st century. *PNAS*, 112(49),
15078-15083.

Case, A. and Deaton, A. (2017). Mortality and morbidity in the 21st
century. *Brookings Papers on Economic Activity*, Spring 2017, 397-476.

Quinones, S. (2015). *Dreamland: The true tale of America's opiate
epidemic*. Bloomsbury Press.

Wu, J. Y. et al. (2020). *COVID-19 US County-level Summaries* (data
repository). <https://github.com/JieYingWu/COVID-19_US_County-level_Summaries>

Shannon M Monnat.(2019) The contributions of socioeconomic and opioid supply factors to us drug mortality
rates: Urban-rural and within-rural differences. Journal of rural studies, 68:319–335

Pamela L Owens, Audrey J Weiss, Marguerite L Barrett, and Lawrence D Reid. (2020) Social determinants of
health and county population rates of opioid-related inpatient stays and emergency department visits.
