---
title: "Lab 7: Measuring latent variables in the social world"
author: "Saaz Kaur Sahdra"
subtitle: 'DATA5207: Data Analysis in the Social Sciences'
output:
  pdf_document:
    toc: TRUE
    toc_depth: 3
    latex_engine: xelatex
  html_document:
    toc: no
    df_print: paged
header-includes:
- \usepackage[x11names]{xcolor}
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
- \usepackage{sectsty}
- \allsectionsfont{\color{ForestGreen}}
mainfont: Avenir Book
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

In Week 7 and 8, you will be working with your groups on the first assessable group exercise. You will have these two labs to finish and upload your work (see the *Group assignment 1* assessment on canvas). Your goal in this project is to come up with a theory that explains quality of life. You are required to work out what might best predict these quality of life scores. 

To do this we will give you a survey and corresponding data (*The World Values Survey*). You will use these data to identify predictors for quality of life. **We will create the dependent variable in this lab**. This will be a measure of quality of life created with a combination of five questions from the survey using factor analysis. These variables are listed below. ***Since they will form part of your dependent variable, they should not be included as predictors for this exercise***. However, if you want to check out whether your predictors have reasonable explanatory power, you can test them using these variables. 

**This week, you will work on developing theories for what might predict higher/lower quality of life.** Think about what drives life quality. Look at questionnaire and codebook. Identify possible predictors. Write up the descriptive analysis (means, sd, or histograms, tables). 

Next week you will test your ideas on the dependent variable we create today. Further details are available below. 

## More details on the group project

This task, which counts towards your group work grade, should not require each group member to spend more than four hours to complete. The three group projects in this semester combined are worth 25% of your total grade for the unit and will be conducted in your groups for the assessable group projects, which should have no more than five members.  
  
In the lecture, we explored the use of different methodologies to understand measures of similarities used in the social sciences (also known as data reduction techniques). These methods are usually adopted in social science research to reduce the number of variables included in a model, or to detect structure in the associations between different variables. In this lab, we will primarily focus on the first of these. This strategy is particularly important for measuring characteristics of individuals for which a single question would often not provide an accurate result; due to measurement error, social desirability bias, or because what we are trying to measure is a latent trait where no one question could possibly capture the true picture. 

The most common method used to do this in the social sciences is factor analysis. While factor analysis expresses the underlying common factors for an entire group of variables, it also helps researchers differentiate these factors by grouping variables into different dimensions or factors, each of which is ideally uncorrelated with the others.

During this lab, we will walk through the process of fitting a factor analysis to survey data on quality of life. We will look at factor analysis as a tool to understand quality of life. This is an appropriate method for this kind of research, as there are a number of variables in our dataset that provide information on the outcome we are interested in, and we can better understand the problem if we decompose the information they contain into a single measure, which we can then study further. The assessable component of the two labs over Week 7 and 8 will require you to undertake an analysis of the measure we create using the other methods you have learned in this subject. 

You should submit your final written document as a *RMarkdown* file through a link that will be provided in this module. **Submit the RMD in a zipped folder with your data**. Make sure your RMD is knittable without any errors. Only one member of your group needs to submit your work. As long as the groups are properly registered on canvas, all members of the group will receive the full grades awarded. 


# Lab 7 Agenda

Your work in this lab consists of two parts. First we will walk you through the use of *factor analysis* in $R$. Then you will be familiarising yourself with the survey data, and examining possible predictors, or independent variables, for this project. 

1. DLD and review
2. Creating the dependent variable 
3. Interpreting factor analysis 
4. Overview of FA
5. Assessable group work Part 1

In next week's lab you will use the output from the factor analysis model we create today to complete your assessable group project. 

\newpage

# DLD

In your DATA5207 local folder, create a new *R Markdown* file and save it with a descriptive name (i.e., Lab7) that will make sense to you when you come back to it in the future.  

Add and edit appropriate YAML headers into the *R Markdown* file. Use the console or other short hand commands to **set your working directory**.  

We start by **loading in our libraries**:

```{r load libraries, message=FALSE, warning=FALSE} 

library(ggplot2)
# install.packages("psych") 
library(psych)
library(tidyverse)

```

*Note:* To run our factor analysis, we use the `psych` package, which we need to download from CRAN using the `install.packages` function before loading the psych package.  
  
We are using the 2018 Australian World Values Survey for this research. This is a survey of Australians, run in conjunction with other similar surveys conducted around the world. It asks respondents more questions than most other surveys, and therefore provides a useful data source for researchers. A copy of the dataset, code book and questionnaire is available in the zipped folder in the module for this class on canvas. These data come from the [Australian Data Archive](https://dataverse.ada.edu.au/dataverse/wvs).

```{r load data, message=FALSE, warning=FALSE}
wvs.dat <- read.csv("wvs_data.csv")
```

## YOUR TASK #1

1. What are latent variables?  
\textcolor{BurntOrange}{
Latent variables are variables that cannot be directly observed or measured but are inferred from other observable variables. In social sciences, concepts like quality of life, happiness, or intelligence are considered latent because they are complex and cannot be captured by a single question. Instead, they are estimated using multiple related survey questions.
}
  
2. What is factor analysis?  
\textcolor{BurntOrange}{
Factor analysis is a statistical method used to identify underlying relationships between variables by grouping them into one or more latent factors. It reduces multiple observed variables into a smaller number of factors, helping to simplify data and measure abstract concepts like quality of life.
}
  
3. What are the assumptions of factor analysis?
\textcolor{BurntOrange}{Factor analysis assumes that the observed variables are linearly related to the underlying latent factors. Key assumptions include: (1) adequate sample size --- a minimum of approximately 100 observations, or at least 5--10 cases per variable; (2) variables should be continuous, or at minimum ordinal with several response options; (3) no severe outliers, as these can distort factor loadings; (4) sufficient intercorrelation among variables, which can be assessed using Bartlett's test of sphericity (tests whether the correlation matrix differs significantly from an identity matrix) and the Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy (values above 0.6 are considered acceptable); and (5) a linear relationship between observed variables and the latent factors.}

## Theory

\textcolor{BurntOrange}{Quality of life is influenced by a combination of personal, social, and economic factors. Individuals with better financial stability, good health, strong social connections, and a sense of personal freedom are more likely to report higher life satisfaction.}

\textcolor{BurntOrange}{Economic theory suggests that higher income and financial security improve access to resources, which enhances well-being. Psychological theories highlight the importance of autonomy and perceived control over life decisions. Additionally, social factors such as trust, relationships, and community engagement can significantly impact happiness and overall life satisfaction.}

\textcolor{BurntOrange}{Based on this, we expect variables related to income, employment status, health, education, and social trust to be strong predictors of quality of life.}

# Creating the dependent varaible 

For this project, you will conduct an analysis on life quality using survey data.

Quality of life is a latent variable. We cannot directly measure it. But, we can use other variables to create a measure for quality of life. Using the survey data we've loaded, we will examine how different personal characteristics might be associated with higher and lower quality of life. We will do this by fitting a factor analysis to five variables which cover reported happiness, life satisfaction, health, and financial situation. These variables are:

- *Q46.* Taking all things together, would you say you are... (variable V10)
- *Q47.* All in all, how would you describe your state of health these days? Would you say it is... (variable V11)
- *Q48.* Some people feel they have completely free choice and control over their lives, while other people feel
that what they do has no real effect on what happens to them. Please use this scale where 1 means "no
choice at all" and 10 means "a great deal of choice" to indicate how much freedom of choice and control
you feel you have over the way your life turns out. (variable V55)
- *Q49.* All things considered, how satisfied are you with your life as a whole these days? (variable V23)
- *Q50.* How satisfied are you with the financial situation of your household? (variable V59)


Using a factor analysis, we will create out dependent variable, a measure for quality of life.

## Standardise variables 

We begin by making sure these are coded the correct way (greater quality of life being the positive score), and we standardise the scores. 

We need to recode our variables. As we do this we provide the variables with more intuitive names, to make our work easier. This can be done with the code:

```{r recode and standardise variables, message=FALSE, warning=FALSE}

wvs.dat <- wvs.dat %>%
  dplyr::mutate(happiness = dplyr::recode(V10, 
                                   '1' = 4,
                                   '2' = 3,
                                   '3' = 2,
                                   '4' = 1, 
                                   '-2' = NULL),
         
         health = dplyr::recode(V11,
                                '1' = 5, 
                                '2' = 4, 
                                '3' = 3,
                                '4' = 2,
                                '5' = 1, 
                                '-2' = NULL),
         
         finances = dplyr::recode(V59, 
                                  '-2' = NULL,
                                  .default = V59),
         
         satisfaction = dplyr::recode(V23, 
                                  '-2' = NULL,
                                  .default = V23),
         
         freedom = dplyr::recode(V55, 
                                  '-2' = NULL,
                                  .default = V55))

```

We use the `recode` function from `dplyr` to both switch around those variables that are coded incorrectly (where the higher value is associated with lower quality of life) and where missing values are included as a numeric value (such as -2 for `V10`), which will result in these values being included in our factor analysis and adding misleading information into our model. Here we are using the `recode()` function from `dplyr`.

## Fitting the factor analysis 

We use the `fa` function, from the `psych` package, to run our factor analysis. More information on this function can be found [here](https://www.rdocumentation.org/packages/psych/versions/1.7.8/topics/fa). We fit the model with this syntax:

\vspace{6mm}

```{r factor analysis}

fa.fit <- fa(wvs.dat[,c("happiness",
                         "health",
                         "finances", 
                         "satisfaction",
                         "freedom")], 
             nfactors=1)

```

Within this function we specify our variables, the number of factors we wish to calculate (in our case, just one). 

It is not shown here, but there are several rotation methods available with factor analysis. These can be characterised as orthogonal, which do not allow the latent dimensions to be correlated, and oblique, which do allow correlation. Varimax is a popular method (especially in the social sciences) for orthogonal rotation in factor and principal components analysis, so in instances where you have multiple dimensions, they will remain uncorrelated. We have not included this here, as we have a single dimensional factor analysis, but good to know for future work you might do where there are two or more factors. If we had wanted to include a Varimax rotation, we could do so by adding `rotate="varimax"` after the `nfactors` command (separated by a comma, of course).  
  
We can take the factor scores from this analysis --- the latent quality of life estimated by our factor analysis --- and save this back into our existing dataset using the code:

```{r grab factor scores}

wvs.dat$life.quality <- as.numeric(fa.fit$scores)

```

This saves the factor scores as a new variable called `life.quality`. We can view a sample of these using the `head` function: 

```{r sample factor scores}

head(wvs.dat$life.quality, 10)

```

# Interpreting factor analysis

The output from your model should look like this: 

```{r factor analysis output, echo=FALSE}

fa.fit

```

There is a lot here, and we do not expect you to understand it all straight away. For this exercise, we will focus on the first table. 

The first column of the top table in our output, labelled `MR1`, shows how each item loads onto the latent trait for quality of life. 'MR' here stands for minimum residual, and its name reflects the fitting method (if we had used a different method we would get a different column name; `ML` for maximum likelihood, for instance). A positive loading means that a higher response on this variable indicates a higher score on the combined measure of life quality created by this analysis. A negative loading (if there was one) would suggest that a higher response predicts lower quality of life. 

It is only this simple to interpret the finding because we recoded our variables. If we had not, we would need to remember the way each variable was coded to infer the meaning of these results. You can see what I mean by replacing `happiness` and `health` with the original variables and re-running the factor analysis. 

Once you have done this, return to the original model. There are three other columns. The second column, `h2`, represents the communalities of the variables. These communalities are the total amount of common variance between the variable and the factor(s). Higher communalities are better. If communalities for a particular variable are low, then that variable may struggle to load significantly on any factor. 

The communalities for the $i$th variable are computed by taking the sum of the squared loadings for that variable. To compute the communality for `happiness`, for instance, we square the factor loading for this variable (if there was more than one dimension, we would square the loading for each then sum the results). This gives us `r round(.64^2, 2)`. You can think of these values as the equivalent of multiple *R^2* values from regression models, predicting the variables of interest for your factors. The communality for a given variable can be interpreted as the proportion of variation in that variable explained by the factor(s). If we perform multiple regression of `happiness` against the factor score, we obtain an *R^2* of `r round(.64^2, 2)`, indicating that about `r 100 * round(.64^2, 2)` per cent of the variation in happiness (as measured by this variable) is explained by the factor score. 

One assessment of how well this model is doing can be obtained from the communalities. What you want to see is values that are close to one. This would indicate that the model explains most of the variation for those variables. In this case, the model does better for some variables than it does for others. Our results suggest the factor analysis does the best job of explaining variation in `satisfaction`. This makes sense, as we're trying to measure quality of life, and self-reported life satisfaction is likely the closest variable to this in the model.

The third column, `u2`, has nothing to do with the band, but is rather the measure of uniqueness. This is the proportion of a variable's variance that is not shared with a factor structure. Unique variance is composed of specific and error variance. The existence of uniquenesses is what distinguishes factor analysis from principal components analysis. If the variables in the model are thought to represent a 'true' or latent part of some phenomena, then factor analysis provides an estimate of the correlations with the latent factor(s) representing the data. If we believe them to be measured without error, then principal components provides the most parsimonious description of the data. When we look at our table factor loading table, the output in the third column is the direct inverse of the second; the more the variance of a item is explained by the factor score(s), the less the proportion a variable's variance will not shared with a factor structure.

The last column, `com`, is the complexity of the factor loadings for that variable (see below). This will be '1' when you only have a single dimension, but will (usually) increase as you increase the number of dimensions in your model.

Just below the first table, we can see that our latent dimension accounted for around 49 per cent of the variance in responses. If we had multiple dimensions we would be provided with a statistic for each dimension and the cumultive variance explained.

## Visualising 

To visualise the loadings of the factor analysis, you can use this code:  
```{r loadings}
fa.diagram(fa.fit)
```

See: https://rstudio-pubs-static.s3.amazonaws.com/498796_05b4e23682d54da387499507495c7ef6.html 
for more details and examples on using fa.  
  
Our created QoL measure is approximately standardised, with a mean of zero and standard deviation of (almost) one. We can then use this to analyse the association between quality of life and different individual characteristics that are also available in this dataset.  
  
This is showing the distribution of quaity of life within the sample.  

```{r histogram_of_quality_of_life, fig.cap = "Distribution of quality of life measure produced by factor analysis.",  fig.width = 5, fig.height = 4, fig.align="center", message = FALSE, warning = FALSE}

ggplot(wvs.dat, aes(life.quality)) + 
  geom_histogram(fill = 'black') + 
  theme_minimal()

```

# Overview of factor analysis steps

1. Data cleaning and recoding
2. Running factor analysis
3. Saving the factor scores
4. Plotting factor scores
5. Interpreting factor analysis
6. Analysing the association between quality of life and other variables (part of your assessable group work!)
7. Assessing the reliability of your model (part of your assessable group work!)

\newpage

# Assessable group work

Now you are ready to start the assessable group exercise. 

## Part 1 (lab 7): Understanding your survey data

In this lab, you will be familiarising yourself with the survey data, and examining possible predictors, or independent variables, for this project. 

As a class, we will first go over creating the dependent variable using factor analysis (above). 

Then, in your groups, you will undertake three main steps:

1. Develop a theory or theories of what factors might influence quality of life, and why. This will inform your selection of independent variables (predictors). This does not need to be sophisticated or complicated. A few simple ideas will suffice. Write this up in the Markdown file you will be submitting.

2. Take the time to familiarise yourself with the survey and data. Using the theory you have developed, select what you believe to be appropriate predictors to explain quality of life from the provided survey data. There should be at least five of these. Examine these predictors in $R$. Look at their mean, standard deviation, distribution, etc. Do they need to be recoded in any way? Write up your data cleaning decisions in the Markdown file, and include any descriptive analysis you do.

Some code to help you get started:  

```{r, results = 'hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(psych)
library(ggpubr)

data <- read.csv("wvs_data.csv")

glimpse(data)

data %>%
  ggpubr::get_summary_stats()
```

3. Finally, properly write up the Introduction, Theory, and Data and Methods section of your report.  

Each step in your Markdown file should be clearly labelled. 

Make sure you SAVE your work so you can continue on it in next week's lab.

\newpage

## Introduction

Australia's population enjoys one of the highest average standards of living in the world, yet quality of life varies substantially across individuals. Understanding what drives these differences is of considerable theoretical and policy interest. This report uses data from the 2018 Australian World Values Survey (WVS) to identify key predictors of individual quality of life. The WVS is a nationally representative survey that covers a broad range of social, economic, and psychological dimensions, making it a particularly valuable resource for this kind of research.

Our dependent variable --- quality of life --- is a latent construct operationalised via factor analysis of five survey items measuring self-reported happiness, health, freedom of choice, life satisfaction, and financial satisfaction. Using this composite measure, we examine six candidate predictors drawn from three theoretical traditions: material deprivation theory, economic resource and anxiety theory, and social capital theory. We hypothesise that individuals who are materially secure, financially stable, and embedded in trusting social networks will report higher quality of life.

\newpage

## Theory

Three theoretical frameworks inform our selection of predictor variables for quality of life.

**Material Deprivation Theory.** A foundational argument in welfare research holds that the inability to meet basic needs --- including access to food, income, and healthcare --- directly undermines physical and psychological well-being (Townsend, 1979). When individuals cannot access essential resources, chronic stress accumulates and the conditions for a satisfying life erode. We include two material deprivation indicators: whether a respondent went without medicine or medical treatment they needed (V190), and whether they went without a cash income (V191). We expect both to be positively associated with quality of life --- those who report *never* experiencing these deprivations should score higher on our QoL measure.

**Economic Resources and Anxiety.** Beyond acute deprivation, broader financial standing and employment security shape life quality. Higher household income expands choice and access to goods and services, while anxiety about losing one's job creates chronic psychological strain that dampens well-being (Marmot, 2004). We include household income group (V237) and concern about job loss (V181) as economic predictors. We expect higher income and lower job anxiety to be associated with higher QoL.

**Social Capital and Safety.** Putnam's (2000) theory of social capital holds that trust in others and participation in community life are important determinants of subjective well-being, as strong social ties provide support and a sense of belonging. Separately, drawing on Maslow's hierarchy of needs, a sense of personal physical security is a prerequisite for higher-order well-being. We include trust in neighbours (V103) and personal sense of security (V170) to capture these dimensions. We expect greater social trust and greater feelings of security to be positively associated with QoL.

\newpage

## Data and Methods

### Data

This analysis draws on the 2018 Australian World Values Survey (WVS Wave 7), which surveyed 1,813 Australians via online and hard-copy questionnaires. The survey covers a broad array of social attitudes, values, and life circumstances. Data were obtained from the Australian Data Archive.

### Dependent Variable

Quality of life is operationalised as a latent factor score derived from a single-factor factor analysis of five items: self-reported happiness (V10), health (V11), freedom of choice (V55), life satisfaction (V23), and financial satisfaction (V59). Responses were recoded so that higher scores indicate better outcomes, and missing values (coded $-2$) were set to missing before analysis. The resulting factor score has a mean of approximately zero and a standard deviation close to one, with higher values indicating higher quality of life.

### Variable Selection

To select predictors in a principled way, we screened a set of theoretically motivated candidate variables by computing the Pearson correlation between each raw variable and the quality of life factor score. Variables used to construct the dependent variable (V10, V11, V55, V23, V59) were excluded from consideration. The six variables with the strongest absolute correlations that also spanned distinct theoretical dimensions were retained.

```{r variable selection plot, fig.width = 6.5, fig.height = 6.5, out.width = "\\textwidth", fig.align = "center", message = FALSE, warning = FALSE, fig.cap = "Absolute correlations between candidate predictors and quality of life. The six selected variables are shown in blue."}

cand_vars <- c("V190", "V191", "V170", "V237", "V103", "V181",
               "V141", "V236", "V104", "V24",  "V57",  "V235",
               "Q56",  "V182", "V240", "V84",  "V249", "V238", "V152")

cand_labels <- c("Gone without medicine (V190)",
                 "Gone without cash income (V191)",
                 "Sense of security (V170)",
                 "Household income (V237)",
                 "Trust in neighbours (V103)",
                 "Worry: job loss (V181)",
                 "Democracy satisfaction (V141)",
                 "Social class (V236)",
                 "Trust in people known (V104)",
                 "Trust in others - general (V24)",
                 "Employment type (V57)",
                 "Family financial trend (V235)",
                 "Living standard vs parents (Q56)",
                 "Worry: children's education (V182)",
                 "Age (V240)",
                 "Interest in politics (V84)",
                 "Employment status (V249)",
                 "Sex (V238)",
                 "Importance of God (V152)")

selected_vars <- c("V190", "V191", "V170", "V237", "V103", "V181")

cor_vals <- sapply(cand_vars, function(v) {
  x <- wvs.dat[[v]]
  x[x == -2] <- NA
  cor(x, wvs.dat$life.quality, use = "complete.obs")
})

cor_df <- data.frame(
  variable  = cand_vars,
  label     = cand_labels,
  abs_corr  = abs(cor_vals),
  selected  = cand_vars %in% selected_vars
)

cor_df$label <- factor(cor_df$label,
                       levels = cor_df$label[order(cor_df$abs_corr)])

ggplot(cor_df, aes(x = abs_corr, y = label, fill = selected)) +
  geom_col() +
  geom_vline(xintercept = min(abs(cor_vals[selected_vars])),
             linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey75"),
                    labels = c("TRUE" = "Selected", "FALSE" = "Not selected"),
                    name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Absolute correlation |r| with quality of life", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank())

```

The six selected variables (shown in blue) all have absolute correlations above the dashed threshold line and together span three distinct theoretical dimensions --- material deprivation, economic security, and social capital --- ensuring that the predictor set is both empirically strong and theoretically diverse.

### Predictor Variables

Six predictor variables are examined, each cleaned by replacing the $-2$ (no answer) code with missing values. Two variables --- sense of personal security (V170) and trust in neighbours (V103) --- are reverse-coded so that higher values uniformly represent better outcomes, consistent with the other predictors.

1. **Household income** (`income`, V237): A 10-point scale from 1 (lowest income group) to 10 (highest income group).
2. **Medical access** (`medical_access`, V190): How often the respondent went without medicine or medical treatment needed (1=Often, 4=Never). Higher scores indicate better access.
3. **Income security** (`income_security`, V191): How often the respondent went without a cash income (1=Often, 4=Never). Higher scores indicate greater income stability.
4. **Personal security** (`security`, V170): Originally coded 1=Very secure to 4=Not at all secure; reverse-coded so that 4=Very secure.
5. **Social trust** (`social_trust`, V103): Originally coded 1=Trust completely to 4=Do not trust at all; reverse-coded so that 4=Trust completely.
6. **Job security** (`job_security`, V181): Concern about losing one's job (1=Very much worried, 4=Not at all worried). Higher scores indicate less concern.

### Recoding

```{r recode predictors, message=FALSE, warning=FALSE}

wvs.dat <- wvs.dat %>%
  dplyr::mutate(

    # Household income: 1 = lowest group, 10 = highest group
    income = dplyr::recode(V237,
                           '-2' = NULL,
                           .default = V237),

    # Medical access: 1 = often without medicine, 4 = never; higher = better
    medical_access = dplyr::recode(V190,
                                   '-2' = NULL,
                                   .default = V190),

    # Income security: 1 = often without cash income, 4 = never; higher = more secure
    income_security = dplyr::recode(V191,
                                    '-2' = NULL,
                                    .default = V191),

    # Personal security: reverse-coded so 4 = very secure, 1 = not at all secure
    security = dplyr::recode(V170,
                              '1' = 4,
                              '2' = 3,
                              '3' = 2,
                              '4' = 1,
                              '-2' = NULL),

    # Social trust in neighbours: reverse-coded so 4 = trust completely, 1 = no trust
    social_trust = dplyr::recode(V103,
                                 '1' = 4,
                                 '2' = 3,
                                 '3' = 2,
                                 '4' = 1,
                                 '-2' = NULL),

    # Job security: 1 = very worried about job loss, 4 = not worried at all
    job_security = dplyr::recode(V181,
                                 '-2' = NULL,
                                 .default = V181)
  )

```

### Descriptive Statistics

```{r descriptive stats, message=FALSE, warning=FALSE}

pred_vars <- c("income", "medical_access", "income_security",
               "security", "social_trust", "job_security")

psych::describe(wvs.dat[, pred_vars])[, c("n", "mean", "sd", "min", "max", "skew", "kurtosis")]

```

The summary table shows that household income has a mean of approximately 5 out of 10, suggesting a roughly symmetric distribution across income groups. Medical access and income security are both right-skewed, with the large majority of respondents reporting they never went without medicine or cash income (values concentrated near 4). Sense of personal security and social trust are similarly concentrated at higher (more positive) values, reflecting a generally secure and socially connected sample. Job security is more evenly spread, indicating a more heterogeneous level of employment-related anxiety in the sample.

### Distributions

```{r plot predictors, fig.width = 6.5, fig.height = 9, out.width = "\\textwidth", fig.align = "center", message = FALSE, warning = FALSE, fig.cap = "Distributions of the six predictor variables."}

p_income <- ggplot(wvs.dat %>% dplyr::filter(!is.na(income)),
                   aes(x = factor(income))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Household Income", x = "Income group (1 = lowest, 10 = highest)", y = "Count") +
  theme_minimal()

p_med <- ggplot(wvs.dat %>% dplyr::filter(!is.na(medical_access)),
                aes(x = factor(medical_access))) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "Often", "2" = "Sometimes",
                               "3" = "Rarely", "4" = "Never")) +
  labs(title = "Medical Access", x = "Gone without medicine", y = "Count") +
  theme_minimal()

p_inc_sec <- ggplot(wvs.dat %>% dplyr::filter(!is.na(income_security)),
                    aes(x = factor(income_security))) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "Often", "2" = "Sometimes",
                               "3" = "Rarely", "4" = "Never")) +
  labs(title = "Income Security", x = "Gone without cash income", y = "Count") +
  theme_minimal()

p_sec <- ggplot(wvs.dat %>% dplyr::filter(!is.na(security)),
                aes(x = factor(security))) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "Not at all\nsecure", "2" = "Not very\nsecure",
                               "3" = "Quite\nsecure", "4" = "Very\nsecure")) +
  labs(title = "Personal Security", x = NULL, y = "Count") +
  theme_minimal()

p_trust <- ggplot(wvs.dat %>% dplyr::filter(!is.na(social_trust)),
                  aes(x = factor(social_trust))) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "No trust\nat all", "2" = "Not much\ntrust",
                               "3" = "Trust\nsomewhat", "4" = "Trust\ncompletely")) +
  labs(title = "Social Trust (Neighbours)", x = NULL, y = "Count") +
  theme_minimal()

p_job <- ggplot(wvs.dat %>% dplyr::filter(!is.na(job_security)),
                aes(x = factor(job_security))) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "Very\nworried", "2" = "A good\ndeal",
                               "3" = "Not\nmuch", "4" = "Not at\nall")) +
  labs(title = "Job Security", x = "Concern about job loss", y = "Count") +
  theme_minimal()

ggpubr::ggarrange(p_income, p_med, p_inc_sec,
                  p_sec, p_trust, p_job,
                  ncol = 2, nrow = 3)

```

The bar charts confirm the patterns observed in the summary statistics. The household income distribution is roughly bell-shaped with a slight left skew, peaking in the mid-to-upper range. Medical access and income security are strongly concentrated at the "Never" category, indicating that material deprivation is relatively uncommon in this Australian sample. Personal security is similarly concentrated at higher values, with over 80 per cent of respondents feeling "quite" or "very" secure. Social trust in neighbours clusters heavily at the "trust somewhat" level (recoded as 3), suggesting a moderate but not unconditional level of community trust. Job security is the most evenly distributed variable, with a notable proportion of respondents expressing significant concern about losing their job or being unable to find one.

